\bf An HMM-based part-of-speech tagger for Turkish

### 2015-12-09 paper/paper.tex
Write some more

\section{Motivation}
\label{sec:motivation}
My main motivation for undertaking this project is to understand the
principles underlying POS-tagging. Due to time and resource constraints, it is
not possible for me to build a state-of-the-art POS-tagger for
Turkish. Nevertheless, I expect my POS-tagger
to work with a decent accuracy that is adequate to be a starting point for
further development. Moreover, \citet{Korkut2015} began the development of a
Turkish NLP library named ``Guguk'', implemented in the programming language Haskell. I intend
to merge this project to Guguk, hence contributing towards
the existence of a Turkish NLP-library in Haskell.

Before proceeding, a few remarks on notational conventions: I use
\texttt{HMM} (with the fixed-width font) to the datatype I'm using from the
\texttt{Data.HMM} package and HMM (with regular font) to refer to the concept.

@@

% Briefly survey the most salient prior work that relates to the problem you wish
% to solve. This section should cite relevant sources. This section should be at
% least one to two paragraphs in length.
\cite{dincer2008suffix} built a POS tagger for Turkish using HMMs on suffixes.
This proves an effective approach seeing that (1) Turkish morphology is
exclusively based on suffixes (not prefixes) (2) morphology is central to the
Turkish language so it is likely that any given word is formed through the
agglutination of at least one morpheme. The tagger described has ``\%90.2
accuracy in the best case''.

@@

I will implement a part-of-speech tagger for the Turkish language. As mentioned
in section~\ref{sec:related_work}, I will mostly focus on parts-of-speech of
words\footnote{I base my definition of a ``word'' on the orthographic space in
  most of the cases. There are certain exceptions to this, such as duplication
  which I denote with the POS tag ``Dup''} and will not be accounting for
morphology, unlike most of the related work I've mentioned. There are definite
problems with this approach. Fixing these would go beyond the scope of this
paper therefore I defer the integration of morphology for further development.

@@

\section{Results}
The resulting program I have implemented is based on bigram hidden markov
models. To summarize it we are making use of three probabilities: (1) $P(t_i |
t_{i-1})$ the probability of being at a state (i.e., a tag) given the previous
state. This is computed\footnote{Although I use the word ``compute'', everything
  is already pre-computed and the referred computation is merely a lookup. This
  is discussed in \ref{par:tag}} by the function \texttt{transFn}. (2) $P(w_i | t_i)$ the
probability of seeing a word having a given tag, computed by the function
\texttt{outFn}. Finally, the probability of a given state being the start state
of a sentence, computed by the function \texttt{outProbs}.

\subsection{A Brief Overview of the Program}
The project consists of two executables: (1) \texttt{create-model} that reads
the data from the corpus and parses it into a form from which we can easily
construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
impossible---to directly serialize \texttt{HMM} since it has three fields
containing functions
\footnote{http://stackoverflow.com/questions/17785916/can-haskell-functions
  -be-serialized for further information.}. We consider this a fault in the
design of the \texttt{Data.HMM} package and urge further development. Although
this has almost no effect on performance, it obfuscates the code.

Since processing the corpus is a resource-demanding operation,
\texttt{create-model} is intended be run only once after which \texttt{run-tag}
can be used to make use of that data.

\subsubsection{\texttt{create-model}}
The executable \texttt{create-model} is implemented in two files:
\texttt{Parse.hs} and \texttt{Main.hs}. \texttt{Parse.hs} handles the parsing of
the corpus.

The function \texttt{freqMap} takes as input a list of orderable elements
(denoted with \texttt{Ord a}) and creates a \texttt{Map} containing, as value,
the frequency of each element.

@@

This function is called several times with lists of values we want to
\emph{count} such as tag bigrams, a given word with a certain tag et cetera.
Pre-computing the counts is almost mandatory since frequencies (i.e., counts)
will need to be accessed frequently when computing the probabilities. Once the
necessary maps are generated, they are written to files in the directory
\texttt{model}.

\subsubsection{\texttt{tag}}
\label{par:tag}
The executable \texttt{run-tag} is where the actual tagging happens. The
frequencies we have pre-computed in \texttt{create-model} are passed in to the
probability function which we give below.
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                   then logFloat $\epsilon$
                                   else (logFloat yCount) / (logFloat xCount)
  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
\end{lstlisting}

This is a straightforward implementation of probability relying on a given count
map. One thing to note is that it returns $\epsilon$ probability (i.e., the
smalles value a \texttt{logFloat}) if a given element is not found.

\texttt{probability} when combined with \texttt{freqMap} forms an
efficient method for implementing Bayes' law. When we are creating the functions
required by \texttt{HMM} we are storing the map returned by a call to
\texttt{freqMap} in the lexical closure of \texttt{probability} by making use
Haskell's currying so that we can eloquently define the two crucial functions as
follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
outFn s e = probability (e, s) wordFreqs taggedWordFreqs
\end{lstlisting}
The probability functions used to create \texttt{transFn} and \texttt{outFn} hold
the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in their
lexical closure, respectively.

Finally we declare a new \texttt{HMM} as follows:

\newpage
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
      newHMM = HMM { states      = possibleTags ∷ [POS]
                     , events      = ws ∷ [String]
                     , initProbs   = initProbFn
                     , transMatrix = transFn
                     , outMatrix   = outFn
                     }
\end{lstlisting}

This is a simple record (a record is like C struct in Haskell) containing the
probability functions we defined.\footnote{For some reason, the author of the
  \texttt{Data.HMM} library named the fields (i.e., attribute), that are of
  function types, using names that end with \texttt{-Matrix} although they are
  \emph{not} matrices. The reader should not be confused by this.}

\subsection{Performance}
Parsing the entire corpus and creating a model out of it takes 10.5 seconds on
average. Tagging a sentence (which we are assuming to have 5 words on average)
takes $~0.3$ seconds. The implementation of the \texttt{Data.HMM} datatype uses
\texttt{Data.Array}
\footnote{https://hackage.haskell.org/package/array-0.5.1.0/docs/Data-Array.html}
which is why it runs reasonably fast.

@@

We have shown that it is possible---though not ideal---to build a POS tagger for
Turkish that is based entirely on a bigram hidden markov model. Let us first
look at the accuracy of this model and propose possible fixes that do not
fundamentally change the structure of the model.

@@

\subsection{Limitations}
There is a fundamental limitation to the POS tagger I built: Turkish morphology
is essential to the overall completeness of Turkish.

In English we entertain the notion of \emph{syntactic correctness}, which is
more or less a formalism of correct ordering of words. A speaker of English that
has synactic competence in the language \citep{bachman1990} may know that a
sentence is well-formed, relying on his knowledge of syntactic rules. However,
for the \textit{words}, knowledge of morphological rules is not adequate to
conclude that a word is made up of correctly ordered morphemes; they have to
rely on their lexical knowledge to verify that the morphological form they
derived/inflected is a valid one. Consider the word ``satisfiableness'' which is
not a valid word according to several dictionaries, although being derived with
valid suffixes. The correct form of this concept is ``satisfiability''.

>>>>

\section{Motivation}
\label{sec:motivation}

The main motivation for this project comes from the lack of a POS tagger for
Turkish exclusively implemented in the programming language Haskell. Moreover, I
believe that implementing a POS tagger from scratch is beneficial in terms of
gain experience with relevant statistical models through this process.

Due to time and resource constraints, it is not plausible to build a
state-of-the-art POS tagger for Turkish. Nevertheless, I expect my POS tagger to
work with decent accuracy that is adequate to be a starting point for further
development. Moreover, \citet{Korkut2015} began the development of a Turkish NLP
library named ``Guguk'', implemented in the programming language Haskell. I
intend to merge this project to Guguk, hence contributing towards the existence
of a Haskell Turkish NLP library; I intend to continue development of this
project as a part of Guguk.

Before I proceed, let me give a few remarks on notational conventions I follow
throughout this text: I use \texttt{HMM} (with the fixed-width font) to the
datatype I'm using from the \texttt{Data.HMM} package and HMM (with regular
font) to refer to the concept.

@@

% Briefly survey the most salient prior work that relates to the problem you wish
% to solve. This section should cite relevant sources. This section should be at
% least one to two paragraphs in length.
\cite{dincer2008suffix} built a POS tagger for Turkish using HMMs on suffixes.
This proves an effective approach seeing that (1) Turkish morphology is
exclusively based on suffixes, in contrast to to prefixes (2) morphology is
central to the Turkish language so it is likely that any given word is formed
through the agglutination of at least severals morphemes. The tagger described
by \citet{dincer2008suffix} has ``\%90.2 accuracy in the best case''.

@@

I will implement a part-of-speech tagger for the Turkish language. As mentioned
in section~\ref{sec:related_work}, I will mostly focus on parts-of-speech of
words\footnote{I base my definition of a ``word'' on the orthographic space in
  most of the cases. There are certain exceptions to this, such as duplication
  which I denote with the POS tag ``Dup''} and will not be accounting for
morphology, unlike most of the related work I've mentioned. There are definite
problems with this approach; fixing these would go beyond the scope of this
paper therefore I defer the integration of morphology for further development.

@@

\section{Results}

The resulting program I have implemented is based on bigram hidden markov model.
To summarize it, I am making use of three probabilities: (1) $P(t_i | t_{i-1})$
the probability of being at a state (i.e., a tag) given the previous state. This
is computed\footnote{Although I use the word ``compute'', everything is already
  pre-computed when a call to this function is made and the referred computation
  is merely a lookup. This is discussed in \ref{par:tag}} by the function
\texttt{transFn}. (2) $P(w_i | t_i)$ the probability of seeing a word with a
given tag, computed by the function \texttt{outFn}. Finally, the probability of
a given state being the start state of a sentence, computed by the function
\texttt{outProbs}.

\subsection{A Brief Overview of the Program}

The project consists of two executables: (1) \texttt{create-model} that reads
the data from the corpus and parses it into a form from which we can easily
construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
impossible---to directly serialize the \texttt{HMM} datatype since it has three
fields containing functions \footnote{See
  http://stackoverflow.com/questions/17785916/can-haskell-functions
  -be-serialized for further information.}. We consider this a fault in the
design of the \texttt{Data.HMM} package and urge further development; this has
almost no effect on performance, but it obfuscates the code.

Since processing the corpus is a resource-demanding operation,
\texttt{create-model} is intended be run only once after which \texttt{run-tag}
can be used to make use of the serialized components of the HMM.

I should note that it is not possible to give a complete account of the
program's logic in this chapter although we explain the most salient components.
The source code is in the \texttt{ayberkt-final_project} repository and it is
well-commented. Should the reader have any concerns regarding implemenetation
they could find it convenient to consult the source code.

\subsubsection{\texttt{create-model}}

The executable \texttt{create-model} is implemented in two files:
\texttt{Parse.hs} and \texttt{Main.hs}. \texttt{Parse.hs} handles the parsing of
the treebank.

The function \texttt{freqMap} found in \texttt{create/Main.hs} takes as input a
list of orderable elements (denoted with \texttt{Ord a}) and creates a
\texttt{Map} containing, as value, the number of ocurrence of each element.

@@

This function is called several times with lists of values we want to
\emph{count} (such as the list of tag bigrams, or the list of all tagged words)
et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
counts) will need to be accessed many times by the viterbi algorithm when
finding the most probable path through the model. Once the necessary maps are
generated, they are written to files in the directory \texttt{model}.

\subsubsection{\texttt{tag}}
\label{par:tag}

The executable \texttt{run-tag} is where the actual tagging happens. The
frequencies we have pre-computed in \texttt{create-model} are passed in to the
probability function which we give below.
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                   then logFloat $\epsilon$
                                   else (logFloat yCount) / (logFloat xCount)
  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
\end{lstlisting}
This is a straightforward implementation of probability relying on a given count
map. One thing to note is that it returns $\epsilon$ probability (i.e., the
smallest value a \texttt{LogFloat}\footnote{A float in the log domain.} can have) if a given element is not found.

\texttt{probability} when combined with \texttt{freqMap} forms an efficient
method for implementing dependent probability.

When we are creating the functions required by \texttt{HMM} we are storing the
map returned by a call to \texttt{freqMap} in the lexical closure of
\texttt{probability} by making use Haskell's currying so that we can eloquently
define the two crucial functions as follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
    transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
    outFn s e = probability (e, s) wordFreqs taggedWordFreqs
\end{lstlisting}
The probability functions used to create \texttt{transFn} and \texttt{outFn}
hold the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in
their lexical closure, respectively.

Finally we declare a new \texttt{HMM} as follows:
\newpage
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
      newHMM = HMM { states      = possibleTags ∷ [POS]
                     , events      = ws ∷ [String]
                     , initProbs   = initProbFn
                     , transMatrix = transFn
                     , outMatrix   = outFn
                     }
\end{lstlisting}

This is a simple record\footnote{A record is a datatype declaration in Haskell
  similar to a struct in C.} containing the probability functions we
defined.\footnote{For some reason, the author of the \texttt{Data.HMM} library
  named the fields (i.e., attribute), that are of function types, using names
  that end with \texttt{-Matrix} although they are \emph{not} matrices. The
  reader should not be confused by this.}

\subsection{Performance}
Parsing the entire corpus and creating a model out of it takes $10.5$ seconds on
average.\footnote{On my i5 MacBook} Tagging a sentence grows with $O(n)$
complexity where $n$ is the number of words in a given sentence. Assuming a
sentence has 5 words on average, this operation takes around $~0.3$ seconds.
\footnote{https://hackage.haskell.org/package/array-0.5.1.0/docs/Data-Array.html}

@@

We have shown that it is possible---though not ideal---to build a POS tagger for
Turkish---a canonical agglutinative language---that is based entirely on a
bigram HMM. In this section I analyze the accuracy of the tagger and discuss
simple improvements I can make, as well as its fundamental limitations fixing
which would require a significant rewrite of the tagger. Let us first look at
the accuracy of this model and propose possible fixes that do not fundamentally
change the structure of the model.

@@

\subsection{Limitations}
There exists a fundamental limitation to the described tagger: Turkish
morphology is essential to the overall completeness of Turkish.

In English we entertain the notion of \emph{syntactic correctness}, which is
more or less a formalism of the correct ordering of words. A speaker of English
that has synactic competence in the language \citep{bachman1990} may know that a
sentence is well-formed, relying on his knowledge of syntactic rules. However,
for the \textit{words}, knowledge of morphological rules is not adequate to
conclude that a word is made up of correctly ordered morphemes; they have to
rely on their lexical knowledge to verify that the morphological form they
derived/inflected is a valid one. Consider the word ``satisfiableness'' which is
not a valid word according to several dictionaries, although being derived with
valid suffixes. The correct form of this concept is ``satisfiability''.
